On the marginal value of swimming in woodland caribou

### 2021-04-12 paper/manuscript.Rmd
cut old ms

Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring
details see Peignier et al. 2019). We identified `r nevents` swimming events over three years (Figure
2). In total, 13 of 29 collared female caribou swam among islands (Figure 2). Of the caribou that
swam, caribou swam on average approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`)
times per year and remained on each island for a median of `r dur` (range = 
`r durMin`-`r durMax`) days before swimming again. We found that caribou residency increased
proportionally with the size of the island, such that caribou spent less time on smaller islands and
more time on larger islands, providing preliminary evidence for links between forage limitation and
the marginal value theorem (Figure 2B).

@@

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Of the caribou that swam, caribou swam on average
approximately `r neventidMean` (range = `r neventidMin`-`r neventidMax`) times
per year and remained on each island for approximately `r dur` (range = 
`r durMin`-`r durMax`) days before swimming again. Based on GPS data to infer
swimming, we separated the population into three apparent groups of individuals:
those that swam regularly between islands (n = 3, every ~`r indMean` days);
those that rarely engaged in swimming events (n = 9, every ~ `r ind2Mean` days);
and those that never swam (n = 17).

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, which have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a few studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitats and foraging
opportunities. We propose that one such explanation, the forage limitation
hypothesis, is the most likely explanation for oceanic swimming for caribou that
live on islands in the Fogo Island archipelago.

Caribou were introduced to Fogo Island between 1964–67 as part of a series of
translocations throughout the province [@Bergerud1989]. Typical caribou
predators such as black bears (*Ursus americanus*) and wolves (*Canis lupus*)
are absent from Fogo Island; the dominant predator is coyote (*Canis latrans*),
which predate calves but it remains unclear what threat they pose, if any, to
adult caribou [@Rayl2014]. Given the minimal risk of predation to adults and low
density of coyote on Fogo Island, we propose that forage limitation is more
likely to drive movement between islands than predator avoidance. For several
decades after their introduction to Fogo Island, the caribou population did not
exceed ~100 individuals [@Bergerud1989, Newfoundland and Labrador Wildlife
Division, unpublished data]. During the 1990s, population density reached ~300
animals and anecdotal evidence suggests caribou were first seen on nearby
islands during this same period. Increased competition for resources on Fogo
Island due to this increase in density could have precipitated movement to other
islands.

Forage limitation and over-grazing is a major concern for caribou populations
and can lead to reduced female reproductive success [@Schaefer2016]. Forage
scarcity has been proposed as a potential reason for caribou moving between
arctic islands on the sea-ice in winter [@Miller1977]. For the Fogo Island
population, as the population grew and predation pressure was insufficient to
maintain population density, forage depletion could lead caribou to make
similar, but riskier and more costly, movements between islands during the
ice-free season. Given the small size of other islands near Fogo (Figure 2),
over-grazing by newly arrived caribou could rapidly deplete forage on an annual
cycle [@Bergerud1990], leading to more frequent movements back and forth between
islands rather than permanent relocation.

The trade-off associated with swimming between islands can be understood as
density-dependent habitat selection governed by the Ideal Free Distribution
[@Morris1987]. Ideal Free Distribution theory predicts a fitness
equilibrium: when the density in a given habitat patch has exceeded the optimum
for fitness within that patch, animals should relocate and settle new habitat
patches, so that fitness is equal across all patches [@Morris1987]. Islands
act as discrete habitat patches in this case, but the costs associated with
swimming create a trade-off when moving between patches. For caribou in the Fogo
Island archipelago, the role of forage limitation in habitat selection patterns
remains unknown, but swimming between islands may be at least partially governed
by density-dependent habitat selection.

Forage limitation may be the ultimate explanation for caribou swimming between
oceanic islands, but a number of proximate mechanisms remain untested. Our
observations of caribou swimming in the ocean lead to five questions for future
research:
<!-- a) For the questions that arise, maybe try to generalize the language a bit
so that they are a bit broader: caribou as model, but emphasize the
ecological/evolutionary process it will comment on. 
b.) Consider tweaking the
intro the the questions less as 'someone could do this' as something along the
lines of : this has ignited in us the idea that we can tailor our data
collection in our closely studies system to access some of these questions. -->

1) How do energetic costs and risks of swimming differ from walking on sea ice?
Walking over ice is less energetically expensive than swimming, but may pose a
greater risk of mortality from falling through the ice. Although the presence
and extent of sea ice varies annually,  we deduce that although the majority of
inter-island movements occur between April 1 and December 31, some inter-island
movements appear to be caribou walking on sea ice between islands (Figure 2B).
Our aim for future research on the trade-offs of these types of movement
considers how the costs, e.g. energy expenditure or mortality, and benefits,
e.g. novel foraging opportunities, may vary seasonally.

<!--2) At which temporal and spatial scales do caribou choose to swim
between islands? Does the time of day, weather, season, or availability
of food influence swimming?-->

<!--3) Does reproductive status (presence of a calf-at-heel) or demographic class
(age or sex) influence propensity for oceanic swimming? Our initial observation
was of an adult male, but all collared individuals were adult females. Are
younger caribou more likely to swim if they are unable to compete with older
individuals for resources? Are males or females without a calf-at-heel more
likely to swim in the summer given they are not restricted based on presence of
a calf-at-heel, as females are? -->

2) If swimming is a function of density-dependent habitat selection, what are
the costs of movement between patches? IFD assumes zero-cost movement, but could
the risk and energetic costs associated with oceanic swimming be estimated using
the potential fitness deficit that caribou are willing to tolerate before
swimming to a new island?

3) In relation to questions 1 and 2: how do individuals vary in their
propensity to swim and their assessment of the costs and benefits of swimming
between islands? We observed some individuals that regularly engaged in swimming
events, while others were never observed swimming in the ocean? Our existing
research on individual behavioural variation in caribou movement ecology
]@Webber2020] provides a basis to investigate the role of inherent individual
differences in swimming behaviour. Specifically, the forage limitation
hypothesis applies uniformly across populations, but individuals could evaluate
the costs and benefits of swimming differently.

We surmise that for Fogo Island, as competition among conspecifics increased
along with population size, density-dependent habitat selection resulted in
expansion of the population to nearby islands. Following Ideal Free Distribution
theory, caribou should swim to new islands when the average fitness of
individuals on the starting island exceeds the density-fitness equilibrium
[@Morris1987]. While this is an ultimate explanation, our data suggest caribou
only remain on smaller islands for a few days at a time, so the fitness
equilibrium remains a theoretical construct. More likely, however, is that
fine-scale competition, density-dependent habitat selection, and forage
depletion drives individuals to periodically swim between islands. Our focal
observations add to the evidence that caribou can, and occasionally do, swim in
the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012]. We suggest that swimming is likely more common than
previously thought for caribou living on oceanic islands, and that forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for this phenomenon. Although our inference is limited to
observations, islands appear to represent discrete foraging patches for
terrestrial animals that can influence fitness via increased foraging
opportunities. We also present several hypotheses related to the nuances and
further variation within the idea that density-dependent habitat selection
governs swimming behaviour, and posit that above and beyond these additional
factors, individuals can vary in their evaluation of costs and benefits of
movement between these habitat patches.

@@

```{r, fig.cap = '(ref:fig1)', fig.width=4}
knitr::include_graphics('../graphics/IMG_1562_crop_edit.jpg')
```

>>>>

Swimming behavior was common among GPS radio-collared caribou in the population (n = 29, for collaring
details see Peignier et al. 2019). We identified `r nevents` swimming events over three years (Figure
2). In total, 13 of 29 collared female caribou swam among islands (Figure 2). Of the caribou that
swam, caribou swam on average approximately `r neventidMedian` (range = `r neventidMin`-` neventidMax`) times per year and remained on each island for a median of `r dur` (range = 
`r durMin`-`r durMax`) days before swimming again. We found that caribou residency increased
proportionally with the size of the island, such that caribou spent less time on smaller islands and
more time on larger islands, providing preliminary evidence for links between forage limitation and
the marginal value theorem (Figure 2B).

@@

@@

```{r, fig.cap = '(ref:fig1)', fig.width=4}
knitr::include_graphics('../graphics/Fig1.jpg')
```
